Grounding the Interaction: Anchoring Situated Discourse in Everyday Human-Robot Interaction

### 2011-10-19 paper_knowledge.tex
Fixed all typos & style mistakes noted by reviewers
The biger questions remain to be dealt with.

A messy table, covered with cardboard boxes, books, video tapes... Thomas is
moving out and is packing everything with the help of Jido, his robot
(Fig.~\ref{fig|vpt}). ``Jido, give me that'', says Thomas while looking at a box that contains a
video tape. The robot smoothly grasps the tape, and hands it to him.

@@

While this kind of interaction should hopefully sound quite familiar in a
foreseeable future, our robots are not yet quite up to the task, neither
regarding natural language understanding nor plan-making and manipulation.
To be combined together, those abilities require an unambiguous and shared
representation of concepts (objects, agents, actions...) underlying the
interaction: what are the prerequisites for such a sentence --``Jido,
give me that''-- to be understood by the robot, correctly interpreted in the
spatial context of the interaction, and eventually transformed into an action?

@@

The next step is to extract and understand the \emph{intent} of the utterance
as thought by the agent. In our example, Thomas obviously wants an action to be
performed by the robot. The action parametrization is conveyed by the semantics
attached to the words and the grammatical structure of the sentence. In our
example, the type of action is given by the verb ``\textit{give}''. Assuming
the robot has some procedural knowledge attached to this symbol, the action
type can be considered as grounded for the robot. We can as well understand
that the recipient of the action is the human, the performer is the robot
itself, and the object acted upon is the tape. The recipient, performer and object are three of the
\emph{thematic roles}~\cite{Gruber1965} that qualify the \emph{give} action. They are necessary to fully ground the sentence\footnote{This analysis has been
inspired on the work of Austin et al.~\cite{Austin1962}, where this type of
sentences correspond to \emph{speech acts}, comprising of \emph{locutionary
act} (the meaning), \emph{illocutionary} (the intent) and possibly
\emph{perlocutionary acts} (implicit speaker's expectation).}.

@@

Besides proposing a new integration model for sensor data, natural language and
symbolic knowledge repositories, our work extends these previous contributions
by tackling more realistic human-robot interactions: open speech; complex,
dynamic, partially unknown human environments; and fully embodied (with arms,
head, ability to move,...) autonomous robots that manipulate a large range of
household objects.

@@

These points highlight some original aspects of a larger cognitive architecture
that has been deployed and tested on several robotic platforms
(including both humanoid robots and service robots), demonstrating the
versatility and hardware-agnosticism of these developments.

@@

Knowledge is represented in ORO in the Description Logics formalism (using the
OWL-DL --\emph{Web Ontology Language - Description Logics}-- language), as
RDF (\emph{Resource Description Framework}) triples (for instance \stmt{robot
isIn kitchen}). We use the Jena\footnote{\url{http://jena.sourceforge.net/}}
framework as the underlying library to load and build in-memory OWL model. We
use it in conjunction with the
Pellet\footnote{\url{http://clarkparsia.com/pellet/}} reasoner to ensure the
continuous classification of the OWL concept graph: at run-time, newly added
statements are continuously reasoned about (\emph{classified}), and at any time,
the server exposes a complete set of statements, both the asserted ones and the
inferred ones. For instance, if \stmt{socrates type Human} and \stmt{Human
subClassOf Mortal} are asserted, the server transparently adds the inferred
statement \stmt{socrates type Mortal}, and a later query retrieving all mortal
entities would return \concept{socrates} as well. The language of the OWL
family use the \emph{Open World Assumption} (if a fact can not be inferred as
true, it does not mean that it is inferred to be false), and the Pellet
reasoner honors this assumption as well.

@@

Next sections describe how symbolic knowledge is actually produced and added to
ORO.

@@

\item \emph{Sees}: An important ability to know about an agent is to predict
``what it can see'', \ie what is within its field of view (FOV). A robot being
able to compute this information can then act accordingly. An example
would be a clarification scenario where the human is searching for an object
and the robot is able to infer that he/she is looking for the one that is not
visible (otherwise the user would not be searching for it).
In figure~\ref{fig::vis} the field of view of a person is illustrated with a
grey cone (broader one). While he is able to see the two small boxes on the
table in front of him, the big box on his right is out of his FOV, and
therefore, he is not able to see it.

@@

The sentence resolution has three steps: {\it(i)} pronouns and anaphora are
replaced by the correct speaker ID and the ID of the last object referred to
(extracted from the dialogue history) respectively, {\it(ii)} nominal groups are
disambiguated and grounded (noun phrase resolution), and {\it(iii)}
verbal groups are resolved and their associated \emph{thematic roles} are
retrieved (verbal phrase resolution).

@@

This second example  also shows the pronouns reference resolution: ``me'' is
replaced by the id of the current speaker while ``you'' is replaced by
\concept{myself} (\concept{myself} always represents the robot itself). When
present, anaphoras (references to previous concepts like ``give me the banana, I
like {\bf it}.'') are also resolved in the same step.

@@

While no examples involving questions have been detailed, \emph{W-} questions
and \emph{yes/no} questions can be processed in a similar way by
\textsc{Dialogs}. For instance, a question like ``What is on the table?''
is grounded (to extract the relation \concept{isOn} and to find what \emph{table}
refers to) and transformed into the following kind of query: \concept{find ?var
[\stmt{?var isOn table1}]}.  Answers are converted back to a full sentence
by the \emph{verbalization} module, and uttered to the human.

@@

This section presents several experiments where the different features of our
approach are represented. The experiments have been conducted on two different platforms: the
\emph{Rosie} manipulator from the Technical Universtiy of Munich, a dual-arm,
holonomic service robot, running the ROS\footnote{\emph{Robotic Operating
System}, \url{http://www.ros.org}} middleware; and the \emph{Jido} robot~\cite{Alami1998a} from LAAS-CNRS, a similar, single-arm, service robot, running the LAAS's {\sc
Genom/pocolibs} stack. Both platforms share the use of ORO, the ontology server proposed in this work.

@@

The \emph{Naming} task uses ORO to anchor perception into the robot's knowledge
through interaction with the user. This task has been implemented on the Rosie
robot at Munich university.

@@

In order to illustrate some of different reasoning abilities of the robot presented in this paper, we have designed the following
daily life situation. Tom and Jerry are moving to London, so they are packing
things in boxes. The scenario takes places in the living-room, where Jido is observing while they move things here and there. To assess the
reasoning abilities of the robot they ask Jido for information (entered through
keyboard). Ideally, the robot should perform actions when required (\eg
hand an object when asking ``give me...''). However, since it is out of the
scope of this work, we do not include any motion from the robot's side. Similar
to the \emph{Spy Game} scenario, perception of objects is done through a tag-based
system and humans are detected through motion capture. We next describe in
detail three situations where we can follow the internal robot's reasoning and
the interaction with the users.

@@

In this paper, we have studied knowledge streams between three components: {\it(1)}
{\sc ORO}, an ontology-based knowledge server that stores and maintains
classified RDF statements produced by other modules in agent-specific
models and allows information to be easily retrieved, either through queries or
via an event system; {\it(2)} {\sc SPARK}, the grounded, human-aware 3D model
of the environment that performs all the spatial reasoning within our
architecture, including reasoning involving motion planing (to compute
reachability of object) and perspective taking, and {\it(3)} {\sc Dialogs}, a
natural language processor that performs simple grammatical parsing of English
language, grounds the semantic content of the utterance (if necessary, also
interacts with the user to disambiguate), and eventually generates a RDF
representation of the sentence.

@@

This architecture moves away from standard layered approaches. Interactions
between components are mostly bidirectional and we do not have a concept of
layers of abstraction (we do, however, have access to the robot lower level
modules to execute actions, but all cognition-related modules live at the same
level). This is especially visible for the dialogue input processing. This
component does not simply act as an alternative perceptual input to the
symbolic database, but also actively queries previously acquired knowledge to
disambiguate and validate the newly created symbolic knowledge.

Regarding the anchoring question, this architecture is
bidirectional. The components we described provide a \textit{bottom-up}
grounding process: SPARK and \textsc{Dialogs} constantly build and push new
symbolic contents about the world to ORO where it becomes accessible to
decisional layers. In parallel, ORO relies on reasoning in a \textit{top-bottom}
way to produce new facts that may trigger in return physical behaviours.

We believe that this \emph{knowledge-oriented} approach has a strong potential
not only to enable rich human-robot interaction, but also as a broader approach
to information alignment and fusion in complex robotic systems.
The versatility of this paradigm could be illustrated by a simple imaginary
scenario with a blind robot and a deaf robot. The blind robot does not see (no
cameras or alike), but someone can verbally describe a scene to it. On the other hand, the
deaf robot has a good vision system, but cannot hear.  Without any
changes to the software architecture described, supervision modules of both robots would be able to perform equally well.

@@

We hope as well that our contribution may contribute to fill the gap between
robotics and psychology. Our architecture provides easier-to-reach entry points
to implement some classical psychology tests to robots. We presented
experiments focused on issues related to perspective taking. By explicitly
enabling independent modeling of the beliefs of each agent, our architecture is
especially well suited to set up cognitive and psychological experiments (such
as the \emph{False-Belief} experiment), which we plan to
further explore.

@@

The main challenge we address in this work can be formulated as \emph{How to
model real-world interaction in a symbolic way, processable by the robot to
make decisions}. In the paper we used several times the term \emph{grounding} to
describe the process of binding percepts to symbols (later organized in a
first-order logic framework).  We would like to relate it to
Sloman~\cite{Sloman2007} stance against the \emph{``Symbol Grounding meme''},
where he argues that symbolic grounding is bound to the representation of
somatic concepts (\ie roughly, the sensori-motor relationships that the robot
learns from its interaction with the world) which in turn severely constraints
the domain of concepts accessible to the robot. We could call this type of
grounding \emph{bottom-up} grounding, and Steels~\cite{Steels2007} claims it is
a solved issue.

@@

Amongst the other shortcomings of our architecture, neither the \emph{domain of
validity} nor the context of a fact are represented in a satisfying way (we do
store some kind of context --the agent mental model for instance--). Those
informations are meta-informations on the knowledge. While the ORO framework allows them through \emph{statement reification}, it does not offer yet
a convenient way to store them. One obvious limitation that derives from
the lack of efficient meta-knowledge is the absence of knowledge history.
With ORO, the robot always lives in the present.

@@

However, we can build a model of the ``human point of view'' because the robot
perceives the human, and is able to estimate, at least partially, what the human
perceives or not. We infer that a human focuses on some object because it points at it, looks at it, and besides, the object is visible to him. This relies on the embodied
nature of the interaction. In turn, this allows us to understand the
meaning of sentences like ``Give me that''.

>>>>

{\em A messy table, covered with cardboard boxes, books, video tapes... Thomas
is moving out and is packing everything with the help of Jido, his robot.
``Jido, give me that'', says Thomas while looking at a box that contains a
video tape. The robot smoothly grasps the tape, and hands it to him.}

@@

While this kind of interaction (Fig.~\ref{fig|vpt}) should hopefully sound
quite familiar in a foreseeable future, our robots are not yet quite up to the
task, neither regarding natural language understanding nor plan-making and
manipulation.  To be combined together, those abilities require an unambiguous
and shared representation of concepts (objects, agents, actions...) underlying
the interaction: what are the prerequisites for such a sentence --``Jido, give
me that''-- to be understood by the robot, correctly interpreted in the spatial
context of the interaction, and eventually transformed into an action?

@@

The next step is to extract and understand the \emph{intended meaning} of the utterance
as thought by the agent. In our example, Thomas obviously wants an action to be
performed by the robot. The action parametrization is conveyed by the semantics
attached to the words and the grammatical structure of the sentence. In our
example, the type of action is given by the verb ``\textit{give}''. Assuming
the robot has some procedural knowledge attached to this symbol, the action
type can be considered as grounded for the robot. We can as well understand
that the recipient of the action is the human, the performer is the robot
itself, and the object acted upon is the tape. The recipient, performer and object are three of the
\emph{thematic roles}~\cite{Gruber1965} that qualify the \emph{give} action. They are necessary to fully ground the sentence\footnote{This analysis has been
inspired on the work of Austin et al.~\cite{Austin1962}, where this type of
sentences correspond to \emph{speech acts}, comprising of \emph{locutionary
act} (the meaning), \emph{illocutionary} (the intent) and possibly
\emph{perlocutionary acts} (implicit speaker's expectation).}.

@@

Besides proposing a new integration model for sensor data, natural language and
symbolic knowledge repositories, our work extends these previous contributions
by tackling more realistic human-robot interactions: less restricted speech
understanding; ability to deal with complex, partially unknown human
environments; and fully embodied (with arms, head,...) autonomous robots that
manipulate a large range of household objects.

@@

These points highlight some original aspects of a larger cognitive architecture
that has been deployed and tested on several mobile robotic platforms
(including both humanoid robots and service robots), demonstrating the
versatility and hardware-agnosticism of these developments.

@@

Knowledge is represented in ORO in the Description Logics formalism (using the
OWL-DL --\emph{Web Ontology Language - Description Logics}-- language), as
RDF (\emph{Resource Description Framework}) triples (for instance \stmt{robot
isIn kitchen}). We use the Jena\footnote{\url{http://jena.sourceforge.net/}}
framework as the underlying library to load and build an in-memory OWL model. We
use it in conjunction with the
Pellet\footnote{\url{http://clarkparsia.com/pellet/}} reasoner to ensure the
continuous classification of the OWL concept graph: at run-time, newly added
statements are continuously reasoned about (\emph{classified}), and at any time,
the server exposes a complete set of statements, both the asserted ones and the
inferred ones. For instance, if \stmt{socrates type Human} and \stmt{Human
subClassOf Mortal} are asserted, the server transparently adds the inferred
statement \stmt{socrates type Mortal}, and a later query retrieving all mortal
entities would return \concept{socrates} as well. The language of the OWL
family make the \emph{Open World Assumption} (if a fact can not be inferred as
true, it does not mean that it is inferred to be false), and the Pellet
reasoner honors this assumption as well.

@@

The next sections describe how symbolic knowledge is actually produced and added to
ORO.

@@

\item \emph{Sees}: An important ability to know about an agent is to predict
``what it can see'', \ie what is within its field of view (FOV). A robot being
able to compute this information can then act accordingly. An example
would be a clarification scenario where the human is searching for an object
and the robot is able to infer that he/she is looking for the one that is not
visible (otherwise the user would not be searching for it).
In Figure~\ref{fig::vis} the field of view of a person is illustrated with a
grey cone (broader one). While he is able to see the two small boxes on the
table in front of him, the big box on his right is out of his FOV, and
therefore, he is not able to see it.

@@

The sentence resolution has three steps: {\it(i)} pronouns and anaphora are
replaced by the correct speaker ID and the ID of the last object referred to
(extracted from the dialogue history) respectively, {\it(ii)} nominal groups are
disambiguated and grounded (noun phrase resolution), and {\it(iii)}
verbal groups are resolved and their associated \emph{thematic roles} are
retrieved (verb phrase resolution).

@@

This second example  also shows the pronoun reference resolution: ``me'' is
replaced by the id of the current speaker, while ``you'' is replaced by
\concept{myself} (\concept{myself} always represents the robot itself). When
present, anaphoras (references to previous concepts like ``give me the banana, I
like {\bf it}.'') are also resolved in the same step.

@@

While no examples involving questions have been detailed, factual \emph{wh-}
questions and polar (\emph{yes/no}) questions can be processed in a similar way
by \textsc{Dialogs}. For instance, a question like ``What is on the table?'' is
grounded (to extract the relation \concept{isOn} and to find what \emph{table}
refers to) and transformed into the following kind of query: \concept{find ?var
[\stmt{?var isOn table1}]}.  Answers are converted back to a full sentence by
the \emph{verbalization} module, and uttered to the human.

@@

This section presents several experiments where the different features of our
approach are represented. The experiments have been conducted on two different
platforms: the \emph{Rosie} manipulator from the Technical Universtiy of
Munich, a dual-arm, holonomic service robot, running the
ROS\footnote{\emph{Robotic Operating System}, \url{http://www.ros.org}}
middleware; and the \emph{Jido} robot~\cite{Alami1998a} from LAAS-CNRS, a
similar, single-arm, service robot, running the LAAS's {\sc Genom/pocolibs}
stack. Both platforms share the use of ORO, the ontology server described in
this work.

@@

The \emph{Naming} task uses ORO to anchor perception into the robot's knowledge
through interaction with the user. This task has been implemented on the Rosie
robot at TU Munich.

@@

In order to illustrate some of different reasoning abilities of the robot presented in this paper, we have designed the following
daily life situation. Tom and Jerry are moving to London, so they are packing
things in boxes. The scenario takes places in the living room, where Jido is observing while they move things here and there. To assess the
reasoning abilities of the robot they ask Jido for information (entered through
keyboard). Ideally, the robot should perform actions when required (\eg
hand an object when asking ``give me...''). However, since it is out of the
scope of this work, we do not include any motion from the robot's side. Similar
to the \emph{Spy Game} scenario, perception of objects is done through a tag-based
system and humans are detected through motion capture. We next describe in
detail three situations where we can follow the internal robot's reasoning and
the interaction with the users.

@@

In this paper, we have studied knowledge streams between three components: {\it(1)}
{\sc ORO}, an ontology-based knowledge server that stores and maintains
classified RDF statements produced by other modules in agent-specific
models and allows information to be easily retrieved, either through queries or
via an event system; {\it(2)} {\sc SPARK}, the grounded, human-aware 3D model
of the environment that performs all the spatial reasoning within our
architecture, including reasoning involving motion planing (to compute
reachability of objects) and perspective taking, and {\it(3)} {\sc Dialogs}, a
natural language processor that performs simple grammatical parsing of English
language, grounds the semantic content of the utterance (if necessary, also
interacts with the user to disambiguate), and eventually generates a RDF
representation of the sentence.

@@

This architecture moves away from standard layered approaches. Interactions
between components are mostly bidirectional and we do not have a concept of
layers of abstraction (we do, however, have access to the lower level modules
of the robot to execute actions, but all cognition-related modules reside at the
same level). This is especially visible for the dialogue input processing. This
component does not simply act as an alternative perceptual input to the
symbolic database, but also actively queries previously acquired knowledge to
disambiguate and validate the newly created symbolic knowledge.

Regarding the anchoring question, this architecture is
bidirectional. The components we described provide a \textit{bottom-up}
grounding process: SPARK and \textsc{Dialogs} constantly build and push new
symbolic contents about the world to ORO where it becomes accessible to
decisional layers. In parallel, ORO relies on reasoning in a \textit{top-down}
way to produce new facts that may trigger in return physical behaviours.

% We believe that this \emph{knowledge-oriented} approach has a strong potential
% not only to enable rich human-robot interaction, but also as a broader approach
% to information alignment and fusion in complex robotic systems.  The
% versatility of this paradigm could be illustrated by a simple imaginary
% scenario with a blind robot and a deaf robot. The blind robot does not see (no
% cameras or alike), but someone can verbally describe a scene to it. On the
% other hand, the deaf robot has a good vision system, but cannot hear.  Without
% any changes to the software architecture described, supervision modules of both
% robots would be able to perform equally well.

@@

% We hope as well that our contribution may contribute to fill the gap between
% robotics and psychology. Our architecture provides easier-to-reach entry points
% to implement some classical psychology tests to robots. We presented
% experiments focused on issues related to perspective taking. By explicitly
% enabling independent modeling of the beliefs of each agent, our architecture is
% especially well suited to set up cognitive and psychological experiments (such
% as the \emph{False-Belief} experiment), which we plan to
% further explore.

@@

The main challenge we address in this work can be formulated as \emph{How to
model real-world interaction in a symbolic way, processable by the robot to
make decisions}. In the paper we used several times the term \emph{grounding} to
describe the process of binding percepts to symbols (later organized in a
first-order logic framework).  We would like to relate it to
Sloman's~\cite{Sloman2007} stance against the \emph{``Symbol Grounding meme''},
where he argues that symbolic grounding is bound to the representation of
somatic concepts (\ie roughly, the sensori-motor relationships that the robot
learns from its interaction with the world) which in turn severely constraints
the domain of concepts accessible to the robot. We could call this type of
grounding \emph{bottom-up} grounding, and Steels~\cite{Steels2007} claims it is
a solved issue.

@@

Amongst the other shortcomings of our architecture, neither the \emph{domain of
validity} nor the context of a fact are represented in a satisfying way (we do
store some kind of context --the agent's mental model for instance--). This
information is meta-information on the knowledge. While the ORO framework
allows them through \emph{statement reification}, it does not offer yet a
convenient way to store them. One obvious limitation that derives from the lack
of efficient meta-knowledge is the absence of knowledge history.  With ORO, the
robot always lives in the present.

@@

However, we can build a model of the ``human point of view'' because the robot
perceives the human, and is able to estimate, at least partially, what the
human perceives or not. We infer that a human focuses on some object because
he/she points at it, looks at it, and besides, the object is visible to him.
This relies on the embodied nature of the interaction. In turn, this allows us
to understand the meaning of sentences like ``Give me that''.
